Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-11-17 grove-paper.tex
fill in section 3, through Graph Decomposition

% XXX "p" is overloaded - could mean pattern or position

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      \textrm{Patterns:}          & p    & {}\in Pat & {}::={} & x \mid \hole                                                            \\
      \textrm{Expressions:\qquad} & e    & {}\in Exp & {}::={} & x \mid \lambda p\mathord{:}\tau.e \mid e~e \mid e + e \mid n \mid \hole \\
      \textrm{Types:}             & \tau & {}\in Typ & {}::={} & \tau \rightarrow \tau \mid Num \mid \hole                               \\
    \end{array}
  \]
  \caption{Syntax as a grammar for the lambda calculus that we are considering.}
  \label{fig:Syntax as a grammar}
\end{figure}

@@

The language we are modeling is shown in \autoref{fig:Syntax as a grammar},
standard for a simply-typed lambda calculus except that all types of terms have
holes (represented by $\hole$). In particular, the argument to a lambda can be
either a bound variable or a hole.

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{ll}
      \arity : \K \rightarrow & \wp(\P)                                        \\
      \hline
      \arity(\Root)={}        & \left\{ \Root \right\}                         \\
      \arity(\PatVar)={}      & \left\{ \right\}                               \\
      \arity(\ExpVar)={}      & \left\{ \right\}                               \\
      \arity(\ExpLam)={}      & \left\{ \LamParam, \LamType, \LamBody \right\} \\
      \arity(\ExpApp)={}      & \left\{ \AppFun, \AppArg \right\}              \\
      \arity(\ExpPlus)={}     & \left\{ \PlusLeft, \PlusRight \right\}         \\
      \arity(\TypArrow)={}    & \left\{ \ArrowArg, \ArrowResult \right\}       \\
      \arity(\TypNum)={}      & \left\{ \right\}                               \\
    \end{array}
  \]
  \caption{Constructors, Indexes and Arity}
  \label{fig:Constructors, Indexes and Arity}
\end{figure}

@@

A graph $\G : \E \rightarrow \Sigma$ is a function from edges to edge states,
where $\E = \U \times \V \times \P \times \V$,
and   unique IDs are drawn from some suitable set $\U$,
and   vertices are drawn from $\V$,
and   positions are drawn from $\P$,
and   edge states are drawn from $\Sigma$, all defined below.

@@

A vertex $v = (u, k)$ identifies a distinct instance of some constructor $k$,
where $k \in \K$ corresponds to a type of expression. Each contructor has an
arity, which we define as a set of positions. A position $p \in \P$ identifies a
location, adjacent to some parent vertex, from which edges to child vertices may
originate. The $\arity$ function in \autoref{fig:Constructors, Indexes and
  Arity} gives the complete mapping from expression constructors to potential edge
positions. Every graph contains a distinguished root vertex, root constructor,
and root position, respectively named $v_0$, $k_0$, and $p_0$.

A graph can be derived from an expression in the calculus by starting with an
empty graph, selecting unique ids $u, u_k$ and a constructor $k$ for the
outermost expression, and creating the edge $\e = (u, v_0, p_0, (u_k, k))$.
Then, for each sub-expression at some position $p \in \arity(k)$, select unique
id $u'$ and constructor $k'$, and create the edge $e' = (u', v, p, v')$, where
$v = (u_k, k)$ and $v' = (u', k')$. Continuing in this manner, treat each
sub-expression, starting with the ones at positions in $\arity(k')$, as if it
were the outermost and create an edge to the vertices representing each of its
immediate sub-expressions. Note that holes are not modeled explicitly in the
graph. When an expression has a hole, it means that the graph has no edges
originating from that position.

@@

% XXX deletes are idempotent

We define graph edit actions as a pair of an edge state, excluding $\bot$, and an edge.
Thus, $\alpha \in \A = \left\{\Plus, \Minus\right\} \times \E$.
As a shorthand, we define $+e$ to be $(\Plus, e)$ and $-e$ to be $(\Minus, e)$.

We define the join operation $s_1 \sqcup s\_2$ to be the least upper bound of $s_1$ and $s_2$ with respect to the $\sqsubset$ ordering.
For example,
$\Plus \sqcup \bot = \Plus$, which means edges can be created if they have never existed,
and $\Plus \sqcup \Minus = \Minus = \Minus \sqcup \Minus$, which means that once an edge is deleted it can never be restored.

%smallest element of $\Sigma$ that is greater than or equal to both $s_1$ and $s_2$.

We define the semantics of graph actions via the following
transition relation between graphs.
\[
  G \overset{\left(s,e\right)}{\longrightarrow} G\left[ e \mapsto s \sqcup G(e) \right\}
\]
In other words, applying graph action $\alpha = \left(s, e\right)$ to a graph $G$
results in $G\left[e \mapsto s \sqcup G(e) \right]$ (i.e., the graph is updated at $e$ to be
the join of $s$ with whatever was already at $e$).

@@

What does a graph decomposition produce? a 4-tuple (reachable, multiparented, deleted, simple cycles).

>>>>

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      \textrm{Operators:}         & op   & {}\in Ops & {}::={} & + \mid * \\
      \textrm{Patterns:}          & q    & {}\in Pat & {}::={} & x \mid \hole \mid \{q|\ldots|q\} \\
      \textrm{Expressions:\qquad} & e    & {}\in Exp & {}::={} & x \mid \lambda q \mathord{:}\tau.e \mid e~e \mid e~op~e \mid n \\
                                  &      &           &         & \phantom{x} \mid \hole \mid \{e|\ldots|e\} \\
      \textrm{Types:}             & \tau & {}\in Typ & {}::={} & \tau \rightarrow \tau \mid Num \mid \hole \mid \{ \tau | \ldots | \tau \} \\
    \end{array}
  \]
  \caption{Syntax as a grammar for the lambda calculus that we are considering.}
  \label{fig:Syntax as a grammar}
\end{figure}

@@

The language we are modeling is shown in \autoref{fig:Syntax as a grammar},
standard for a simply-typed lambda calculus except that all types of terms have
unfilled holes (represented by $\hole$), and most types of terms have conflict holes (of the form $\{\cdot | \ldots | \cdot\}$). In particular, the argument to a lambda can be either a bound variable or a hole.

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{ll}
      \arity : \K \hfil \rightarrow & \wp(\P)                                        \\
      \hline
      \arity(k_0)={}        & \left\{ \Root \right\}                         \\
      \arity(x \in Pat)={}      & \left\{ \right\}                               \\
      \arity(x \in Exp)={}      & \left\{ \right\}                               \\
      \arity(\lambda q \mathord{:}\tau.e)={}      & \left\{ \mathtt{Param}, \mathtt{Type}, \mathtt{Body} \right\} \\
      \arity(e~e)={}      & \left\{ \mathtt{Fun}, \mathtt{Arg} \right\}              \\
      \arity(e~op~e)={}     & \left\{ \mathtt{L}, \mathtt{R} \right\}         \\
      \arity(n)={}      & \left\{ \right\}                               \\
      \arity(\tau \rightarrow \tau)={}    & \left\{ \mathtt{Arg}, \mathtt{Result} \right\}       \\
      \arity(Num)={}      & \left\{ \right\}                               \\
    \end{array}
  \]
  \caption{Constructors, Indexes and Arity}
  \label{fig:Constructors, Indexes and Arity}
\end{figure}

@@

A graph $\G : \E \rightarrow \Sigma$ is a function from edges to edge states,
where $\E = \U \times \V \times \P \times \V$,
unique IDs are drawn from some suitable set $\U$,
vertices are drawn from $\V$,
positions are drawn from $\P$,
and edge states are drawn from $\Sigma$, all defined below.

@@

A vertex $v = (u, k)$ identifies a distinct instance of some constructor $k$,
where $k \in \K$ corresponds to one of the forms defined in \autoref{fig:Syntax
as a grammar} that is not an operator or a hole. Each contructor has an arity,
which we define as a set of positions. A position $p \in \P$ identifies a
location, adjacent to some parent vertex, from which edges to child vertices may
originate. The $\arity$ function in \autoref{fig:Constructors, Indexes and
Arity} gives the complete mapping from constructors to potential edge
positions. Every graph contains a distinguished root vertex, root constructor,
and root position, respectively named $v_0$, $k_0$, and $p_0$.

A graph can be derived from an expression in the calculus by starting with an
empty graph, selecting unique ids $u, u_k$ and constructor $k$ for the outermost
expression, and creating edge $\e = (u, v_0, p_0, (u_k, k))$. Then, for each
sub-term at some position $p \in \arity(k)$, select unique id $u'$ and
constructor $k'$ and create edge $e' = (u', v, p, v')$ with $v = (u_k, k)$, $v'
= (u', k')$. Continuing in this manner, treat each sub-term, starting with the
ones at positions in $\arity(k')$, as if it were the outermost and create edges
to the vertices representing its immediate sub-terms until only operators and
holes remain. Note that empty holes are not modeled explicitly in the graph.
Wherever an empty hole term appears, the graph has no edges originating from
that position.

@@

We define a graph edit $\alpha \in \A = \left\{\Plus, \Minus\right\} \times \E$
as a pair of an edge state, excluding $\bot$, and an edge. As a shorthand, we
define $+\e$ to be $(\Plus, \e)$ and $-\e$ to be $(\Minus, \e)$.

We define the join operation $s_1 \sqcup s_2$ to be the least upper bound of
$s_1$ and $s_2$ with respect to the $\sqsubset$ ordering. Accordingly, $\Plus
\sqcup \bot = \Plus$ means that edges can only be created if they do not already
exist, and $\Plus \sqcup \Minus = \Minus = \Minus \sqcup \Minus$ means that once
an edge is deleted it can never be restored.

We define the semantics of graph actions via the following transition relation
between graphs.
\[
  G \overset{(s,\e)}{\longrightarrow} G\left[ \e \mapsto s \sqcup G(\e) \right\}
\]
In words, applying graph action $\left(s, \e\right)$ to graph $G$ results in the
updated graph $G' = G\left[\e \mapsto s \sqcup G(\e) \right]$, wherein the edge
state associated with edge $\e$ in $G$ becomes the join of $s$ with the state of
$\e$ in $G$, and $G'(\e') = G(\e')$ for all other $\e' \ne \e$ in the domain of
$G$.

@@

If we translate the graph into expressions in the calculus, interactions with
the graph can be modeled as a stream of user actions as described in
\autoref{sub:User Actions}. Our decomposition algorithm partitions the graph
into disjount connected components. We encode cycles by resolving all but one
vertex of each multi-edge into a by-id reference to the one vertex.

The result of decomposition is a 4-tuple $(e, MP, D, SC) \in Exp \times
\wp({Exp}) \times \wp({Exp}) \times \wp({Exp})$. Expression $e$ covers the
portion of the graph reachable from the root vertex $v_0$. $MP$ contains the
expressions that involve vertices with multiple parents. $D$ contains deleted
expressions, i.e., those not reachable from $v_0$, that do not contain cycles.
$SC$ contains the deleted expressions that do not involve multiple parents but
do contain cycles.
